Testing and Isolation Efficacy: Insights from a Simple Epidemic Model

### 2021-02-23 SIR_manuscript.tex
updated methods

In order to understand the effect of testing processes on an epidemic dynamics, we deveopled a mechanistic SIR-type model with epidemic and testing components. Here, we focus on the the effect of testing intensity, different levels of testing ``focus'' (from random to highly targeted), and rate of test return on the epidemic dynamics. 
Our model provides insights to the sensitivity of the epidemiological dynamics, through $\Rnum$, with respect to the undelying epidemic and testing parameters.
\david{This paragraph does not flow from the rest of the intro, and is a weird way to end when the point of the paper is about testing.}\ali{major changes done}

@@

We developed a deterministic model, Eqs. \eqref{model}, which groups individuals based on disease status and testing status. Disease states include Susceptible, Infectious and Recovered (thus this is an SIR-type model), and testing status categorizes people as \emph{untested}, waiting-for-\emph{positive}, waiting-for-\emph{negative}, or \emph{confirmed positive} (Figure~\ref{fig:flowchart}). Symbolically, the testing status of an individual in the disease compartment $X$, where $X \in \{S,I,R\}$ , is reflected in the subscript, namely $X\_u$, $X\_p$, $X\_n$ and $X\_c$, for \emph{untested}, waiting-for-\emph{positive}, waiting-for-\emph{negative}, or \emph{confirmed positive}, respectively. Further, two `accumulator' compartments, $N$ and $P$, were also incorporated in the model in order to collect cumulative reported negative or positive tests. The model and details of calculation of the basic reproduction number $\Rnum$ are presented in the appendixs.

\begin{figure}[!h] 
\begin{center} 
\includegraphics[scale=0.3]{./pix/sir_comp.png}
\caption{\small Flowchart of the SIR (Susceptible-Infectious-Recovered) model, \ref{model}. For the description of the parameters see \cref{tab:params}. \todo{change the subs to roman font}
\label{fig:flowchart}}
\end{center} 
\end{figure}

@@

In order to link to more applied models\david{what does that mean?  what is an ``applied model''?}, we constructed this model so we could specify the total \percap testing rate. We do this by defining the weighted size of the testing pool $W = W_S S\_u + W_I I\_u + W_R R\_u$, and calculating a scaling parameter for testing as:
\begin{equation}
\label{sigma}
\sigma = \frac{\rho N_0}{W}.
\end{equation}
Thus, the \percap testing rate for compartment $X$ is $F_X=\sigma W_X$, where $X \in \{S,I,R\}$. 
For a high-sensitivity test, infected people typically flow through to the ``confirmed positive'' ($I\_c$, $R\_c$) compartments and are thus unavailable for further testing.  Over the course of the epidemic, a fixed testing rate as specified in \eqref{sigma} can (if large enough) exhaust the pool of people available for testing, leading to a singularity when no one is left untested.  Although this phenomenon does not affect our analysis of $\Rnum$, it can affect the temporal dynamics (we discuss an adjustment to the model that solves this problem in the appendix).

\david{A comment on notation:  You haven't actually defined $S_u$, $I_u$ etc.  In any case, I urge you to use macros and use Roman font for subscripts that are textual as opposed to variables.  So $S_{\rm u}$ and $I_{\rm p}$ etc.  Then it is clear the subscript isn't a variable with a value.}\ali{done, $p_S$ as is?}

The classical SIR model is based on the following implicit assumptions; well-mixed population, homogeneity of the population\david{What's the difference btw ``well-mixed'' and ``homoegeneous''?}\ali{ non structured eg age, ok?}, exponentially distributed duration of infection and large population size (see, e.g., \cite{keeling2011modeling}). In addition to these standard assumptions, our model, \ref{model}, assumes: (i) there is a single force of infection, $\Lambda$, defined as follows
\begin{equation}
\label{Lambda}
\Lambda=\beta \frac{(I\_u+\eta\_w I\_n+\eta\_w I\_p+ \eta\_c I\_c)}{N_0},
\end{equation}
across all susceptible pools (see Table~\ref{tab:params} for details\david{there are no details in the table; you need to do a better job of explaining what's going on}\ali{help?}), (ii) $\eta\_c \leq \eta\_w$, i.e., the individuals awaiting test results have a higher transmission probability than the reported individuals. For this analysis, we also assume a perfectly specific test ($p_S=0$). This last assumption combined with the assumption that no individual is in waiting-for-\emph{positive} or \emph{confirmed positive} compartments, i.e., $S\_p(0)=S\_c(0)=0$, reduces the model to 10 equations (equations c and d  in \eqref{model} are eliminated).

@@

\david{Details such as the fact that you used ggplot are not important to state in the main text.  On the other hand, there isn't a clear narrative here.  The reader should feel that a story is being told.}\ali{partially addressed}

>>>>

In order to understand the effect of testing processes on an epidemic dynamics, we deveopled a mechanistic SIR-type model with epidemic and testing components. Here, we focus on the the effect of testing intensity, different levels of testing ``focus'' (from random to highly targeted), and rate of test return on the epidemic dynamics. 
Our model provides insights to the sensitivity of the epidemiological dynamics, through $\Rnum$, with respect to the undelying epidemic and testing parameters.
\david{This paragraph does not flow from the rest of the intro, and is a weird way to end when the point of the paper is about testing.}\ali{some edits/rewording are done, ok?}

@@

We developed a deterministic model, Eqs. \eqref{model}, which groups individuals based on disease status and testing status. Disease states include Susceptible, Infectious and Recovered (thus this is an SIR-type model), and testing status categorizes people as \emph{untested}, waiting-for-\emph{positive}, waiting-for-\emph{negative}, or \emph{confirmed positive} (Figure~\ref{fig:flowchart}). Symbolically, the testing status of an individual in the disease compartment $X$, where $X \in \{S,I,R\}$, is reflected in the subscript, namely $X\_u$, $X\_p$, $X\_n$ and $X\_c$, for \emph{untested}, waiting-for-\emph{positive}, waiting-for-\emph{negative}, or \emph{confirmed positive}, respectively. Further, two `accumulator' compartments, $N$ and $P$, were also incorporated in the model in order to collect cumulative reported negative or positive tests. The model and details of calculation of the basic reproduction number $\Rnum$ are presented in the appendixs.

\begin{figure}[!h] 
\begin{center} 
\includegraphics[scale=0.3]{./pix/sir_comp.png}
\caption{\small Flowchart of the SIR (Susceptible-Infectious-Recovered) model, \ref{model}. Here, the disease-based status of a compartment $X$, where $X \in \{S,I,R\}$, is combined with the testing-based status including $X\_u$, $X\_p$, $X\_n$ and $X\_c$, for \emph{untested}, waiting-for-\emph{positive}, waiting-for-\emph{negative}, or \emph{confirmed positive}, respectively. Also,  $\Lambda$ is the force of infection with definition in Eq.~\eqref{Lambda}, $\gamma$ is the recovery rate, $\omega$ is the rate of test return, $F_X$ and $p_X$ represent the \percap testing rate and the probability of positive tests, respectively, for compartment $X$. For further description of the parameters see \cref{tab:params}. \ali{(i) will change the subs to roman font, (ii) not sure if ref color is helpful in u compartment?}
\label{fig:flowchart}}
\end{center} 
\end{figure}

@@

In order to link to more applied models, i.e., including both epidemiological and testing processes, we constructed this model so we could specify the total \percap testing rate. We do this by defining the weighted size of the testing pool $W = W_S S\_u + W_I I\_u + W_R R\_u$, and calculating a scaling parameter for testing as:
\begin{equation}
\label{sigma}
\sigma = \frac{\rho N_0}{W}.
\end{equation}
Thus, the \percap testing rate for compartment $X$ is $F_X=\sigma W_X$, where $X \in \{S,I,R\}$. 
For a high-sensitivity test, infected people typically flow through to the ``confirmed positive'' ($I\_c$, $R\_c$) compartments and are thus unavailable for further testing.  Over the course of the epidemic, a fixed testing rate as specified in \eqref{sigma} can (if large enough) exhaust the pool of people available for testing, leading to a singularity when no one is left untested.  Although this phenomenon does not affect our analysis of $\Rnum$, it can affect the temporal dynamics (we discuss an adjustment to the model that solves this problem in the appendix).

The classical SIR model is based on the following implicit assumptions; well-mixed population, homogeneity of the population (i.e., all individuals are equaly susciptable and equaly infectious for the same length of time when infected), exponentially distributed duration of infection and large population size (see, e.g., \cite{keeling2011modeling}). In addition to these standard assumptions, our model, \ref{model}, assumes: (i) there is a single force of infection (new cases per unit time), $\Lambda$, defined as follows
\begin{equation}
\label{Lambda}
\Lambda=\beta \frac{(I\_u+\eta\_w I\_n+\eta\_w I\_p+ \eta\_c I\_c)}{N_0},
\end{equation}
across all susceptible pools with transmission rate $\beta$ and relative probability of transmission for ``waiting'' and \emph{confirmed positive} individuals, $\eta\_w$ and $\eta\_c$ respectively, (see Table~\ref{tab:params} for further details), (ii) $\eta\_c \leq \eta\_w$, i.e., the individuals awaiting test results have a higher transmission probability than the reported individuals. For this analysis, we also assume a perfectly specific test ($p_S=0$). This last assumption combined with the assumption that no individual is in waiting-for-\emph{positive} or \emph{confirmed positive} compartments, i.e., $S\_p(0)=S\_c(0)=0$, reduces the model to 10 equations (equations c and d  in \eqref{model} are eliminated).

@@

\david{Details such as the fact that you used ggplot are not important to state in the main text.  On the other hand, there isn't a clear narrative here.  The reader should feel that a story is being told.}\ali{partially addressed?, feedback is welcomed}
